### 2015-06-06 chapters/ipa_meets_unicode.tex
tried a different itemize formatting

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
problems with SAMPA are that (i) it is only a partial encoding of the IPA and
(ii) it encodes different languages in separate data tables, instead of a
universal alphabet, like IPA. SAMPA tables are derived from phonemes appearing
in several European languages that were developed as part of a European
Commission-funded project to address technical problems like electronic mail
exchange (what is now simply called email). SAMPA is essentially a hack to work
around displaying IPA characters, but it provided speech technology and other
fields a basis that has been widely adopted and used in code. So, SAMPA was a
collection of tables to be compared, instead of a large universal table
representing all languages. An extended version of SAMPA, called X-SAMPA, set
out to include every symbol in the IPA chart including all diacritics
\citep{WellsND}. X-SAMPA was considered more universally applicable because it
consisted of one table that encoded the set of characters that represented
phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
for speech technology and as an encoding system in computational linguistics.
Eventually, ASCII-encoding of the IPA became deprecated through the advent of
the Unicode Standard. Note however that many popular software packages used for
linguistic analyses still require ASCII input, e.g.~RuG/L04 and
SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and
\url{http://www.splitstree.org/}, respectively}

There are a few pitfalls to be aware of when using the Unicode Standard
to encode IPA. As we have said before, from a linguistic perspective it might
look like the Unicode Consortium is making incomprehensible decisions, but it is
important to realize that the consortium has tried and is continuing to try to
be as consistent as possible across a wide range of use cases, and it does place
linguistic traditions above other orthographic possibilities. In general, we
strongly suggest linguists not to complain about any decisions in the Unicode
Standard, but to try and understand the rationale of the Unicode Consortium
(which in our experience is almost always well-conceived) and devise ways to
work with any unexpected behavior. Many of the current problems derive from the
fact that the IPA is clearly historically based on the Latin script, but
different enough from most other Latin-based writing systems to warrant special
attention. This ambivalent status of the IPA glyphs (partly Latin, partly
special) is unfortunately also attested in the treatment of IPA in the Unicode
Standard. In retrospect, it might have been better to consider the IPA (and
other transcription systems) to be a special or new kind of script within the
Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we will have to learn to deal with the specific pitfalls of IPA within
the Unicode Standard. In this section, we will describe these pitfalls in some
detail.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 13 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see \cite[Appendix C]{Moran2012}.}

@@

Furthermore, even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}. Conversely,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use <a> as
long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'' at:
\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}

@@

It is not just the Unicode Standard that offers multiple options for encoding
the IPA. Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable for interoperability between resources written in IPA. We
consider it crucial to distinguish between ``lax'' IPA, for which it is
sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are 
aware of the following double options in the IPA, which will be discussed in 
turn below:

@@

\

\noindent The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
for high tone.} The IPA also proposes a less widespread used, but more
consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
five different levels, and sequences of these letters can be used to indicate
contours. Well-designed fonts will even merge a sequence of tone letters into a
contour. For example, compare the font Linux Libertine, which does not merge
tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
For strict IPA encoding we propose to standardize on tone letters.

@@

Likewise, the tie bar to indicate affricates, doubly articulated consonants or
diphthongs can be placed either below or on top of the base characters, e.g.
<\charis{t͡s}> or <\charis{t͜s}>. For strict IPA encoding we propose to standardize on the tie bar
above the base characters, using Unicode \textsc{combining double inverted
breve} at \uni{0361}. Note that this Unicode character is placed between the two 
base characters to be combined. In principle, it is possible to combine more 
than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
necessary, we consider this possible, even though the rendering will never look 
good.

@@

\begin{itemize} 
	\item <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} 
	  (use <dz> instead) 
    \item <ʤ> \textsc{latin small letter dezh digraph} at \uni{02A4}
      (use <dʒ> instead)
    \item <ʥ> \textsc{latin small letter dz digraph with curl} at \uni{02A5}
      (use <dʑ>)
    \item <ʦ> \textsc{latin small letter ts digraph} at \uni{02A6} 
      (use <ts> instead)
	\item <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} 
	  (use <tʃ> instead) 
    \item <ʨ> \textsc{latin small letter tc digraph with curl} at \uni{02A8}
      (use <tɕ>)
   	\item <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9}
	  (use <fŋ> instead) 
\end{itemize}

@@

\begin{itemize}
	\item <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
	\item <œ> \textsc{latin small ligature oe} at \uni{0153} 
	\item <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
	\item <æ> \textsc{latin small letter ae} at \uni{00E6} 
\end{itemize}

@@

Although many combinations of base character with diacritic are treated as 
canonical equivalent with precomposed characters, there are a few combinations 
in IPA that allow for multiple, apparently identical, encodings that are not 
canonical equivalent. The following elements should not be treated as diacritics 
when encoding IPA in Unicode:
\begin{itemize}
  \item <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
  \item <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
  \item <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
  \item <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
  \item <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
  \item <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
\end{itemize}

@@

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if such a decomposition is not intended, because it is meaningless in IPA.

@@

\begin{itemize}
	\item[] Length marks, namely 
	\begin{itemize}
	  \item[] <ː> \textsc{modifier letter triangular colon} at \uni{02D0}
	  \item[] <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}
	\end{itemize}

	\item[] Tone letters, like 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˧> \textsc{modifier letter mid tone bar} at \uni{02E7}
	  \item[] and others like this
	\end{itemize}

	\item[] Superscript letters, like 
	\begin{itemize}
	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] and many more like this
	\end{itemize}

	\item[] the rhotic hook <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
\end{itemize}

@@

In accordance with the absence of any specification of ordering in the IPA, the
Unicode Standard likewise does not propose any standard orderings. Both leave it
to the user to be consistent. However, there is one aspect of ordering for which
the Unicode Standard does present a canonical solution, which is uncontroversial
from a linguistic perspective. Diacritics in the Unicode Standard
(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
Combining Classes. In practice, the diacritics are distinguished by their
position relative to the base character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values for a
detailed description}.} When applying a Unicode normalization (NFC or NFD, see
previous section), the diacritics in different positions are put in a specified
order. This process therefore harmonizes the difference between different
encodings, for example, of a midtone creaky voice vowel <ḛ̄>. This grapheme
cluster can be encoded either as <e> + <̄> + <̰> or as <e> + <̰> + <̄> . To
prevent this twofold encoding, the Unicode Standard specifies the second
ordering as canonical (diacritics below before diacritics above).

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization.
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
<ç>) will be decomposed. In general, Unicode decomposition does not behave like
a feature decomposition as expected from a linguistic perspective. It is thus
important to consider Unicode decomposition only as a technical procedure, and
not assume that it is linguistically sensible.

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions.\footnote{The most recent version of these conventions is online:
\url{http://phoible.github.io/conventions/}} The conventions are influenced by
the linguistic literature, though some ad-hoc decisions had to be taken. The
goal was to explicitly define all character sequences so that the vast variety
of phonemes found in descriptions of the world's language were normalized into
consistent character sequences, e.g.~if one language description uses and
another , when both are intended to be phonetically equivalent, then a decision
to normalize to one form was taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Spacing Modifier Letters ordering:} ( unreleased <̚> \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>) → ( palatalized <ʲ>) → ( labialized <ʷ>) → ( velarized <ˠ>) → ( pharyngealized <ˤ>) → ( aspirated <ʰ> \textbar{} ejective <ʼ>) → ( long <ː> \textbar{} half-long <ˑ>) 
\end{itemize}

@@

\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Combining Diacritical Marks (below) ordering:} ( dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) → ( fronted <u̟> \textbar{} backed <e̠>) →( lowered <e̞> \textbar{} raised <e̝>) → ( ATR <e̘ e̙>) → ( voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>) → ( syllabic <n̩> \textbar{} non-syllabic <e̯>) 
\end{itemize}

@@

\begin{itemize}
	\itemsep1pt\parskip0pt\parsep0pt 
	\item \textsc{Combining Diacritical Marks (above) ordering:} (centralized <ë> \textbar{} mid-centralized <e̽>) → (extra short <ĕ>) →( tone accents, e.g. <è> ) → ( Spacing Modifier Letters ) → ( tone letters, e.g. <e˦>) 
\end{itemize}

>>>>

To alleviate this problem, during the late 1980s, SAMPA (Speech Assessment
Methods Phonetic Alphabet) was created to represent IPA symbols with 7-bit
printable ASCII character sequences, e.g. <p\textbackslash> for [ɸ]. Two
problems with SAMPA are that (i) it is only a partial encoding of the IPA and
(ii) it encodes different languages in separate data tables, instead of a
universal alphabet, like IPA.\@ SAMPA tables are derived from phonemes appearing
in several European languages that were developed as part of a European
Commission-funded project to address technical problems like electronic mail
exchange (what is now simply called email). SAMPA is essentially a hack to work
around displaying IPA characters, but it provided speech technology and other
fields a basis that has been widely adopted and used in code. So, SAMPA was a
collection of tables to be compared, instead of a large universal table
representing all languages. An extended version of SAMPA, called X-SAMPA, set
out to include every symbol in the IPA chart including all diacritics
\citep{WellsND}. X-SAMPA was considered more universally applicable because it
consisted of one table that encoded the set of characters that represented
phones/segments in IPA across languages. SAMPA and X-SAMPA have been widely used
for speech technology and as an encoding system in computational linguistics.
Eventually, ASCII-encoding of the IPA became deprecated through the advent of
the Unicode Standard. Note however that many popular software packages used for
linguistic analyses still require ASCII input, e.g.~RuG/L04 and
SplitsTree4.\footnote{See \url{http://www.let.rug.nl/kleiweg/L04/} and
\url{http://www.splitstree.org/}, respectively}

There are a few pitfalls to be aware of when using the Unicode Standard
to encode IPA.\@ As we have said before, from a linguistic perspective it might
look like the Unicode Consortium is making incomprehensible decisions, but it is
important to realize that the consortium has tried and is continuing to try to
be as consistent as possible across a wide range of use cases, and it does place
linguistic traditions above other orthographic possibilities. In general, we
strongly suggest linguists not to complain about any decisions in the Unicode
Standard, but to try and understand the rationale of the Unicode Consortium
(which in our experience is almost always well-conceived) and devise ways to
work with any unexpected behavior. Many of the current problems derive from the
fact that the IPA is clearly historically based on the Latin script, but
different enough from most other Latin-based writing systems to warrant special
attention. This ambivalent status of the IPA glyphs (partly Latin, partly
special) is unfortunately also attested in the treatment of IPA in the Unicode
Standard. In retrospect, it might have been better to consider the IPA (and
other transcription systems) to be a special or new kind of script within the
Unicode Standard, and treat the obvious similarity to Latin glyphs as a
historical relic. All IPA glyphs would then have their own code points, instead
of the current situation in which some IPA glyphs have special code points,
while others are treated as being identical to the regular Latin characters.
Yet, the current situation, however unfortunate, is unlikely to change, so as
linguists we will have to learn to deal with the specific pitfalls of IPA within
the Unicode Standard. In this section, we will describe these pitfalls in some
detail.

@@

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 13 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}.}

@@

Furthermore, even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}. Conversely,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as [+back] results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
[+back] vowels), as opposed to /i/ and /e/.''} They simply use <a> as
long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'' at:
\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}

@@

It is not just the Unicode Standard that offers multiple options for encoding
the IPA.\@ Even the IPA specification itself offers some flexibility in how
transcriptions have to be encoded. There are a few cases in which the IPA
explicitly allows for different options of transcribing the same phonetic
content. This is understandable from a transcriber's point of view, but it is
not acceptable for interoperability between resources written in IPA.\@ We
consider it crucial to distinguish between ``lax'' IPA, for which it is
sufficient that any phonetically-trained reader is able to understand the
transcription, and ``strict'' IPA, which should be standardized on a single
unique encoding for each sound, so search will work across resources. We are 
aware of the following double options in the IPA, which will be discussed in 
turn below:

@@

The first case in which the IPA allows for different encodings is the question
of how to transcribe tone. There is an old tradition to use diacritics on vowels
to mark different tone levels, e.g. <ȅèée̋>.\footnote{There are at least two
different Unicode homoglyphs for the low and high level tones, namely <\ ̀~>
\textsc{combining grave tone mark} at \unif{0340} vs.\@ <\ ̀~> \textsc{combining
grave accent} at \unif{0300} for low tone, and <\ ́~> \textsc{combining acute tone
mark} at \unif{0341} vs.\@ <\ ́~> \textsc{combining acute accent} at \unif{0301}
for high tone.} The IPA also proposes a less widespread used, but more
consistent option of tone letters, e.g. <˥˦˧˨˩>. Tone letters in the IPA have
five different levels, and sequences of these letters can be used to indicate
contours. Well-designed fonts will even merge a sequence of tone letters into a
contour. For example, compare the font Linux Libertine, which does not merge
tone letters <{\fontspec{LinLibertineO}˥˨˧˩}>, with the font CharisSIL, which
merges this sequence of four tone letters into a single contour <\charis{˥˨˧˩}>.
For strict IPA encoding we propose to standardize on tone letters.

@@

Finally, the tie bar to indicate affricates, doubly articulated consonants or
diphthongs can be placed either below or on top of the base characters, e.g.
<\charis{t͡s}> or <\charis{t͜s}>. For strict IPA encoding we propose to standardize on the tie bar
above the base characters, using Unicode \textsc{combining double inverted
breve} at \uni{0361}. Note that this Unicode character is placed between the two 
base characters to be combined. In principle, it is possible to combine more 
than two base characters by repeating the tie bar, like in <a͡o͡u>. If really
necessary, we consider this possible, even though the rendering will never look 
good.

@@

\begin{itemize} 
	\item[] <ʣ> \textsc{latin small letter dz digraph} at \uni{02A3} 
	  (use <dz>) 
    \item[] <ʤ> \textsc{latin small letter dezh digraph} at \uni{02A4}
      (use <dʒ>)
    \item[] <ʥ> \textsc{latin small letter dz digraph with curl} at \uni{02A5}
      (use <dʑ>)
    \item[] <ʦ> \textsc{latin small letter ts digraph} at \uni{02A6} 
      (use <ts>)
	\item[] <ʧ> \textsc{latin small letter tesh digraph} at \uni{02A7} 
	  (use <tʃ>) 
    \item[] <ʨ> \textsc{latin small letter tc digraph with curl} at \uni{02A8}
      (use <tɕ>)
   	\item[] <ʩ> \textsc{latin small letter feng digraph} at \uni{02A9}
	  (use <fŋ>) 
\end{itemize}

@@

\begin{itemize}
	\item[] <ɮ> \textsc{latin small letter lezh} at \uni{026E} 
	\item[] <œ> \textsc{latin small ligature oe} at \uni{0153} 
	\item[] <ɶ> \textsc{latin letter small capital oe} at \uni{0276} 
	\item[] <æ> \textsc{latin small letter ae} at \uni{00E6} 
\end{itemize}

@@

Although many combinations of base character with diacritic are treated as 
canonical equivalent with precomposed characters, there are a few combinations 
in IPA that allow for multiple, apparently identical, encodings that are not 
canonical equivalent. The following elements should not be treated as diacritics 
when encoding IPA in Unicode:
\begin{itemize}
  \item[] <\ {\large  ̡}\ > \textsc{combining palatalized hook below} at \uni{0321}
  \item[] <\ \ {\large  ̢}> \textsc{combining retroflex hook below} at \uni{0322}
  \item[] <\ \ {\large  ̵}> \textsc{combining short stroke overlay} at \uni{0335}
  \item[] <\ \ {\large  ̷}> \textsc{combining short solidus overlay} at \uni{0337}
  \item[] <\ \ {\large  ̴}> \textsc{combining tilde overlay} at \uni{0334}
  \item[] <{\large ˞}> \textsc{modifier letter rhotic hook} at \uni{02DE}
\end{itemize}

@@

Reversely, <ç> \textsc{latin small letter c with cedilla} at \uni{00E7} will be
decomposed into <c> \textsc{latin small letter c} at \uni{0063} with 
<\ \ {\large  ̧}> \textsc{combining cedilla} at \uni{0327}, 
also if such a decomposition is not intended, because it is meaningless in 
IPA.\@

@@

\begin{itemize}

	\item[] Length marks, namely: 
	\begin{itemize}
	  \item[] <ː> \textsc{modifier letter triangular colon} at \uni{02D0}
	  \item[] <ˑ> \textsc{modifier letter half triangular colon} at \uni{02D1}
	\end{itemize}

	\item[] Tone letters, like: 
	\begin{itemize} 
	  \item[] <˥> \textsc{modifier letter extra-high tone bar} at \uni{02E5}
	  \item[] <˧> \textsc{modifier letter mid tone bar} at \uni{02E7}
	  \item[] and others like this
	\end{itemize}

	\item[] Superscript letters, like:
	\begin{itemize}
	  \item[] <ʰ> \textsc{modifier letter small h} at \uni{02B0}
	  \item[] <ˤ> \textsc{modifier letter small reversed glottal stop} at \uni{02E4}
	  \item[] and many more like this
	\end{itemize}

	\item[] The rhotic hook:
	\begin{itemize}
	  \item[] <˞> \textsc{modifier letter rhotic hook} at \uni{02DE}
	\end{itemize}

\end{itemize}

@@

\subsubsection*{Canonical combining classes}

In accordance with the absence of any specification of ordering in the IPA, the
Unicode Standard likewise does not propose any standard orderings. Both leave it
to the user to be consistent. However, there is one aspect of ordering for which
the Unicode Standard does present a canonical solution, which is uncontroversial
from a linguistic perspective. Diacritics in the Unicode Standard
(i.e.~Combining Diacritical Marks, see above) are classified in Canonical
Combining Classes. In practice, the diacritics are distinguished by their
position relative to the base character.\footnote{See
\url{http://unicode.org/reports/tr44/\#Canonical\_Combining\_Class\_Values} for a
detailed description.} When applying a Unicode normalization (NFC or NFD, see
Section~\ref{pitfall-canonical-equivalence}), the diacritics in different
positions are put in a specified order. This process therefore harmonizes the
difference between different encodings, for example, of a midtone creaky voice
vowel <ḛ̄>. This grapheme cluster can be encoded either as <e> + <\ \ ̄> + <\ \
̰> or as <e> + <\ \ ̰> + <\ \ ̄>. To prevent this twofold encoding, the Unicode
Standard specifies the second ordering as canonical (in this case, diacritics
below are put before diacritics above).

When encoding a string according to the Unicode Standard, it is possible to do
this either using the NFC (composition) or NFD (decomposition) normalization.
Decomposition implies that precomposed characters (like <á> \textsc{latin small
letter a with acute} at \uni{00E1}) will be split into its parts. This might
sound preferable for a linguistic analysis, as the different diacritics are
separated from the base characters. However, note that most attached elements
like strokes (e.g.~in the <ɨ>), retroflex hooks (e.g.~in <ʐ>) or rhotic hooks
(e.g.~in <ɝ>) will not be decomposed, but strangely enough a cedilla (like in
<ç>) will be decomposed (see Section~\ref{pitfall-missing-decomposition}). In
general, Unicode decomposition does not behave like a feature decomposition as
expected from a linguistic perspective. It is thus important to consider Unicode
decomposition only as a technical procedure, and not assume that it is
linguistically sensible.

\subsubsection*{Diacritic ordering}

Facing the problem of specifying a consistent ordering of diacritics while
developing a large database of phonological inventories from the world's
languages, \citet[540]{Moran2012} defines a set of diacritic ordering
conventions.\footnote{The most recent version of these conventions is online at
\url{http://phoible.github.io/conventions/}} The conventions are influenced by
the linguistic literature, though some ad-hoc decisions had to be taken. The
goal was to explicitly define all character sequences so that the vast variety
of phonemes found in descriptions of the world's language were normalized into
consistent character sequences, e.g.~if one language description uses and
another, when both are intended to be phonetically equivalent, then a decision
to normalize to one form was taken. For example, when a character sequence
contains more than one character in Spacing Modifier Letters, the order that is
proposed is the following (where <\textbar{}> indicates \textit{or} and <→>
indicates \textit{precedes}):

\begin{itemize}
	\item[] \textsc{Spacing Modifier Letters ordering:}
	\begin{itemize}
	  \item[→] (unreleased <\ \ ̚ > \textbar{} lateral release <ˡ> \textbar{} nasal release <ⁿ>)
	  \item[→] (palatalized <ʲ>)
	  \item[→] (labialized <ʷ>)
	  \item[→] (velarized <ˠ>)
	  \item[→] (pharyngealized <ˤ>)
	  \item[→] (aspirated <ʰ> \textbar{} ejective <ʼ>)
	  \item[→] (long <ː> \textbar{} half-long <ˑ>)
	\end{itemize}
\end{itemize}

@@

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (below) ordering:}
	\begin{itemize}	
	  \item[→] (dental <t̪> \textbar{} laminal <t̻> \textbar{} apical <t̺>) 
	  \item[→] (fronted <u̟> \textbar{} backed <e̠>)
	  \item[→] (lowered <e̞> \textbar{} raised <e̝>) 
	  \item[→] (ATR <e̘ e̙>)
	  \item[→] (voiced <s̬> \textbar{} voiceless <n̥> \textbar{} creaky voice <b̰> \textbar{} breathy voice <b̤>)
	  \item[→] (syllabic <n̩> \textbar{} non-syllabic <e̯>)
	\end{itemize}
 \end{itemize}

@@

\begin{itemize}
	\item[] \textsc{Combining Diacritical Marks (above) ordering:}
	\begin{itemize}
	  \item[→] (centralized <ë> \textbar{} mid-centralized <e̽>)
	  \item[→] (extra short <ĕ>)
	  \item[→] (tone accents, e.g. <è>)
	  \item[→] (Spacing Modifier Letters, see above)
	  \item[→] (tone letters, e.g. <˦>)
	\end{itemize}	 
\end{itemize}
